Particle Smoothing Algorithms for Variable Rate Models

### 2013-01-15 13-01-14_final_submission/VRsmoothing.tex
Final submitted version

\documentclass[10pt,twocolumn,twoside]{IEEEtran}

@@

\graphicspath{{./images/}{./images/finance/}{./images/tracking/}}
%\interdisplaylinepenalty=1000

@@

\author{Pete~Bunch*,~\IEEEmembership{Member,~IEEE,}
        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{P. Bunch and S. Godsill are with the Department
of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
}%\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}

@@

This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo (MCMC) algorithms \cite{Green1995}.% or in multiple target tracking scenarios \cite{Vo2005}.

@@

This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,% (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
%
%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A x_{n-1} + w_n \\
 y_n & = & C x_{n} + v_n
\end{IEEEeqnarray}

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1}^-)$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-)} \nonumber \\
  &=& S(\tau_{K_n},u_{K_n}, t_n) / S(\tau_{K_{n-1}},u_{K_{n-1}}, t_{n-1}) \nonumber \\
  & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

%For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}). For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that it cannot occur before $t_{n-1}$,
%%
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
%  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
%\end{IEEEeqnarray}

@@

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling an entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula, % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}     .
\end{IEEEeqnarray}

%The choice of proposal weights, $\{v_{n-1}^{(i)}\}$, requires particular attention in the design of VRPFs. In some models a changepoint may not have an immediate effect on the observations, especially if a jump occurs in some quantity which is only observed via its integral, e.g. if there is a jump in the acceleration of a moving object, yet only the position is measured, the change will not be apparent until several more observations have been made. In the meantime, particles which contain a changepoint at the correct time may all have been removed by the resampling process. To avoid this loss of good particles, proposal weights should be chosen which preserve a significant number of low-weight particles. One scheme which has been found to work well is described in \cite{Godsill2007}, in which proposal weights are given by:

@@

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity; because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}.% Since these algorithms are based on the FFBS algorithm, the convergence results of \cite{Godsill2004,Douc2011} are believed to apply.

@@

The filter and filter-smoother were run using using $\aleph_F = 1000$ particles, while the smoother used $\aleph_F = 100$ particles and was used to simulate $\aleph_S = 100$ smoothed paths. These values lead to roughly equal running time for the smoother and filter-smoother. Bootstrap proposals were used for the filter. An example forward realisation simulated from the model is shown in Fig.~\ref{fig:example_data}. The sampled trend trajectories produced by the filter-smoother and VRPS are shown in Fig.~\ref{fig:example_state} (the equivalent graphs for value are not particularly informative as this quantity is observed, and the errors produced by both algorithms are small). Notice that the path diversity is dramatically improved by the VRPS.
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{example_data.pdf}
\caption{An example simulated data set, showing value (top) and trend (bottom). Value observations are overlayed. Jump times are shown as dotted vertical lines.}
\label{fig:example_data}
\end{figure}
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_filtersmoother_state.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_smoother_state.pdf}}
\caption{Trend estimates from (a) the filter-smoother and (b) smoother. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
\label{fig:example_state}
\end{figure}
%
Performance was evaluated by testing on 100 scenarios, each with 1000 observations. At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle. The resulting mean absolute error over all scenarios is shown in Fig.~\ref{fig:finance_rmse}, demonstrating a small improvement in accuracy from the smoother. Mean TNEES is shown in Fig.~\ref{fig:finance_mtnees}, showing a similar efficiency between filter and smoother. The filter-smoother produces a mean TNEES close to $1$ at most time instants, because there is often only one unique particle in the set.% are displayed in table~\ref{tab:finance_state_performance}
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{finance_value_rmse.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{finance_trend_rmse.pdf}}
\caption{Value (a) and trend (b) mean absolute error over 100 scenarios using filter, filter-smoother, and smoother.}
\label{fig:finance_rmse}
\end{figure}
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{finance_mtnees.pdf}
\caption{Mean TNEES over 100 scenarios using filter, filter-smoother, and smoother. Optimum value: 0.54}
\label{fig:finance_mtnees}
\end{figure}
%
%\begin{table}%
%\caption{State estimation performance.}
%\label{tab:finance_state_performance}
%\centering
%\renewcommand{\arraystretch}{1.5}
%\begin{tabular}{|c|c|c|c|}
%\hline
% &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
%\hline \hline
%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
%\hline
%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
%\hline
%\end{tabular}
%\end{table}
%
%The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}. This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint and state distributions.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=0.75\columnwidth]{finance_unique_particles.pdf}
%\caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
%\label{fig:finance_unique_particles}
%\end{figure}
%
As well as providing more accurate and more diverse estimates, the smoother is also capable of preserving multi-modality in its representation of the posterior distribution, which is lost by the filter smoother due to the particle collapse to a single value. This is highlighted in Fig.~\ref{fig:finance_multimodality}, which shows the trend estimates from a segment of simulated data containing a small jump. In the filter-smoother results, none of the particles have a jump at this point. In contrast, a significant number of the smoother particles contain a jump around about the right place. This ability to represent multi-modality is a key advantage of good particle smoothing algorithms.
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.45\columnwidth]{finance_filter_multimodality.pdf}}
\subfloat[]{\includegraphics[width=0.45\columnwidth]{finance_smoother_multimodality.pdf}}
\caption{Filter-smoother (a) and smoother (b) estimates of trend around a jump, highlighting the ability of the smoother to represent multi-modal distributions.}
\label{fig:finance_multimodality}
\end{figure}

@@

An example trajectory is shown in Fig.~\ref{fig:simulated_trajectory}, with the results from the filter-smoother and smoother in Fig.\ref{fig:2D_particle_results}. The improved particle diversity achieved by the smoothing algorithm is readily apparent.
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{simulated_problem.pdf}
\caption{An example simulated data set, showing true position (dashed line), observations (dots) and changepoint positions (crosses). }
\label{fig:simulated_trajectory}
\end{figure}
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{2DFilter.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{2DSmoother.pdf}}
\caption{Filter-smoother (a) and smoother (b) particle estimates of position (solid lines), with particle changepoint positions (crosses). }
\label{fig:2D_particle_results}
\end{figure}
%
Performance was evaluated by testing on 100 independently generated scenarios, each with 500 observations. At each observation time, an estimate of the position and velocity is made by taking the average of values from the array of particles. The resulting mean absolute error over all scenarios is shown in Fig.~\ref{fig:tracking_rmse}, again demonstrating a small improvement in accuracy from the smoother. The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the filter-smoother. Mean TNEES is displayed in Fig.~\ref{fig:tracking_mtnees}. It shows that while the smoother does not match the efficiency of the filter, it is a significant improvement over the filter-smoother, which has a Mean TNEES of $1$ at almost all time instants, because there is only one unique particle in the set. %in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{tracking_pos_rmse.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{tracking_vel_rmse.pdf}}
\caption{Position (a) and velocity (b) mean absolute error over 100 scenarios using filter, filter-smoother, and smoother.}
\label{fig:tracking_rmse}
\end{figure}
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{tracking_mtnees.pdf}
\caption{Mean TNEES over 100 scenarios using filter, filter-smoother, and smoother. Optimum value: 0.73}
\label{fig:tracking_mtnees}
\end{figure}
%
%The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:tracking_unique_particles}. This illustrates again the principal advantage of the VRPS algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariances.
%
%\begin{table}%
%\caption{State estimation performance.}
%\label{tab:tracking_state_performance}
%\centering
%\renewcommand{\arraystretch}{1.5}
%\begin{tabular}{|c|c|c|c|}
%\hline
% &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
%% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
%\hline \hline
%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean position estimate RMSE}} & 2.14 & 1.36 & 1.12 \\
%\hline
%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean velocity estimate RMSE}} & 1.89 & 1.03 & 0.85 \\
%\hline
%\end{tabular}
%\end{table}
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=0.75\columnwidth]{tracking_unique_particles.pdf}
%\caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
%\label{fig:tracking_unique_particles}
%\end{figure}
%
Finally, in Fig.~\ref{fig:varying_num_pts} the variation in accuracy and efficiency of the smoothing particle approximation is demonstrated as the number of smoothing particles varies. As expected, the rate of improvement diminishes and becomes negligible as the number of smoothing particles exceeds the number of filtering particles.
%
\begin{figure}
\centering
\subfloat[]{\includegraphics[width=0.45\columnwidth]{pos_RMSE_vs_numpts.pdf}}
\subfloat[]{\includegraphics[width=0.45\columnwidth]{MTNEES_vs_numpts.pdf}}
\caption{Average position error (a) and Mean TNEES (b) over 100 scenarios with varying number of smoother particles and 50 filter particles.}
\label{fig:varying_num_pts}
\end{figure}

@@

\appendices
%\section{The Backwards Kalman Filter} \label{app:backwards_KF}
%
%A backwards Kalman filter \cite{Fraser1969,Kitagawa1994} is used to calculate the joint likelihood term, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$, when the state dynamics are linear-Gaussian. Assume that at time $t_{n+1}$ there is a Gaussian density such that,
%%
%\begin{equation}
% p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber    .
%\end{equation}
%
%The state transition model is linear-Gaussian, so the backwards predictive density is given by,
%%
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
%  & = & \int p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+)  p(x_{n+1}|x_n, \tilde{\theta}_{n}^+) dx_{n+1} \nonumber \\
%  & = & \int \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
%  &   & \qquad \mathcal{N}(x_{n+1}|A_{n+1} x_{n}, Q_{n+1}) dx_{n+1} \nonumber \\
%  & = & \int \tilde{Z}_{n+1} ||A_{n+1}^{-1}|| \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
%  &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
%  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \nonumber
%\end{IEEEeqnarray}
%
%where,
%%
%\begin{IEEEeqnarray}{rCl}
% \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
% \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \nonumber \\
% \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1} \nonumber   .
%\end{IEEEeqnarray}
%
%$||A||$ denotes the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
%
%The update step simply involves multiplying by the likelihood of the observation at time $t_n$,
%%
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
% & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
% & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
% & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n) \nonumber     ,
%\end{IEEEeqnarray}
%
%where,
%%
%\begin{IEEEeqnarray}{rCl}
% \tilde{\mu}_n & = & C_n \tilde{m}_n^-  \nonumber \\
% \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n  \nonumber \\
% \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
% \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
% \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \nonumber \\
% \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n) \nonumber   .
%\end{IEEEeqnarray}
%
%All steps use only standard Gaussian identities.

%\section{Initialisation of the Backwards Kalman Filter} \label{app:init_backwards_KF}
%
%The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model using the method of \cite{Kitagawa1994}. By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state. This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
%%
%\begin{IEEEeqnarray}{rCl}
% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
% & = & \sqrt{\frac{|V_{N-L}|}{|\Gamma_{N-L}|}} \mathcal{N}(x_{N-L} | \nu_{N-L}, V_{N-L} ) \nonumber
%\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
% V_{N-L} & = & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} \nonumber \\
% \nu_{N-L} & = & V_{N-L} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N} \nonumber     ,
%\end{IEEEeqnarray}
%
%where $\mathbf{y}_{N-L:N}$ is a vector of concatenated observations, and $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models,
%%
%\begin{IEEEeqnarray}{rCl}
%\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} \nonumber \\
%%\end{IEEEeqnarray}
%%\begin{IEEEeqnarray}{rCl}
%H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix} \nonumber \\
%%\end{IEEEeqnarray}
%%\begin{IEEEeqnarray}{rCl}
% \Gamma_{N-1} & = & K_{N-L} \begin{bmatrix} Q_N & 0 & \dots & 0 \\ 0 & Q_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & Q_{N-L+1} \end{bmatrix} K_{N-L}^T \nonumber \\
%  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix} \nonumber \\
%%\end{IEEEeqnarray}
%%\begin{IEEEeqnarray}{rCl}
% K_{N-L} & = & \begin{bmatrix} C & CA & \dots & CA^{L-1} \\ 0 & C & \dots & CA^{L-2} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & C \\ 0 & 0 & \dots & 0 \end{bmatrix} \nonumber
%\end{IEEEeqnarray}
%
%In the above, it is assumed for clarity that $A_n = A$ and $C_n = C$ are constant matrices. This assumption is easily relaxed. This inversion should be effected for the smallest value of $L$ such that the matrix inverses exist.

@@

%\newpage
\bibliographystyle{IEEEtran}
\bibliography{D:/pb404/Dropbox/PhD/Cleanbib}

%\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Bunch.tex}
%\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Godsill.tex}

>>>>

%\documentclass[10pt,twocolumn,twoside]{IEEEtran}
\documentclass[peerreview,11pt,draftcls,onecolumn]{IEEEtran}

@@

\graphicspath{{./images/}{./images/finance/}{./images/tracking/}}

@@

\author{Pete~Bunch*,~\IEEEmembership{Member,~IEEE,}
        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{Copyright (c) 2012 IEEE. Personal use of this material is permitted. However, permission to use this material for any other purposes must be obtained from the IEEE by sending a request to pubs-permissions@ieee.org.}%
\thanks{P. Bunch and S. Godsill are with the Department
of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
}%\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}

@@

This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo (MCMC) algorithms \cite{Green1995}.

@@

This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A x_{n-1} + w_n \\
 y_n & = & C x_{n} + v_n
\end{IEEEeqnarray}

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1}^-)$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-)} \nonumber \\
  &=& S(\tau_{K_n},u_{K_n}, t_n) / S(\tau_{K_{n-1}},u_{K_{n-1}}, t_{n-1}) \nonumber \\
  & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
\end{IEEEeqnarray}

@@

For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}). This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) until one falls after the current time, $t_n$. This final future changepoint is discarded. (This process can be thought of as sampling an entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula,
%
\begin{IEEEeqnarray}{rCl}
w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\theta_{n}^-, y_{1:n-1}) \label{eq:bootstrap_vrpf_weights}     .
\end{IEEEeqnarray}

@@

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity; because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}.

@@

The filter and filter-smoother were run using using $\aleph_F = 1000$ particles, while the smoother used $\aleph_F = 100$ particles and was used to simulate $\aleph_S = 100$ smoothed paths. These values lead to roughly equal running time for the smoother and filter-smoother. Bootstrap proposals were used for the filter. An example forward realisation simulated from the model is shown in Fig.~\ref{fig:example_data}. The sampled trend trajectories produced by the filter-smoother and VRPS are shown in Fig.~\ref{fig:example_state} (the equivalent graphs for value are not particularly informative as this quantity is observed, and the errors produced by both algorithms are small). Notice that the path diversity is dramatically improved by the VRPS.
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{example_data.pdf}
\caption{An example simulated data set, showing value (top) and trend (bottom). Value observations are overlayed. Jump times are shown as dotted vertical lines.}
\label{fig:example_data}
\end{figure}
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_filtersmoother_state.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_smoother_state.pdf}}
\caption{Trend estimates from (a) the filter-smoother and (b) smoother. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
\label{fig:example_state}
\end{figure}
%
Performance was evaluated by testing on 100 scenarios, each with 1000 observations. At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle. The resulting mean absolute error over all scenarios is shown in Fig.~\ref{fig:finance_rmse}, demonstrating a small improvement in accuracy from the smoother. Mean TNEES is shown in Fig.~\ref{fig:finance_mtnees}, showing a similar efficiency between filter and smoother. The filter-smoother produces a mean TNEES close to $1$ at most time instants, because there is often only one unique particle in the set.
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{finance_value_rmse.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{finance_trend_rmse.pdf}}
\caption{Value (a) and trend (b) mean absolute error over 100 scenarios using filter, filter-smoother, and smoother.}
\label{fig:finance_rmse}
\end{figure}
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{finance_mtnees.pdf}
\caption{Mean TNEES over 100 scenarios using filter, filter-smoother, and smoother. Optimum value: 0.54}
\label{fig:finance_mtnees}
\end{figure}
%
As well as providing more accurate and more diverse estimates, the smoother is also capable of preserving multi-modality in its representation of the posterior distribution, which is lost by the filter smoother due to the particle collapse to a single value. This is highlighted in Fig.~\ref{fig:finance_multimodality}, which shows the trend estimates from a segment of simulated data containing a small jump. In the filter-smoother results, none of the particles have a jump at this point. In contrast, a significant number of the smoother particles contain a jump around about the right place. This ability to represent multi-modality is a key advantage of good particle smoothing algorithms.
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.45\columnwidth]{finance_filter_multimodality.pdf}}
\subfloat[]{\includegraphics[width=0.45\columnwidth]{finance_smoother_multimodality.pdf}}
\caption{Filter-smoother (a) and smoother (b) estimates of trend around a jump, highlighting the ability of the smoother to represent multi-modal distributions.}
\label{fig:finance_multimodality}
\end{figure}

@@

An example trajectory is shown in Fig.~\ref{fig:simulated_trajectory}, with the results from the filter-smoother and smoother in Fig.\ref{fig:2D_particle_results}. The improved particle diversity achieved by the smoothing algorithm is readily apparent.
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{simulated_problem.pdf}
\caption{An example simulated data set, showing true position (dashed line), observations (dots) and changepoint positions (crosses). }
\label{fig:simulated_trajectory}
\end{figure}
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{2DFilter.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{2DSmoother.pdf}}
\caption{Filter-smoother (a) and smoother (b) particle estimates of position (solid lines), with particle changepoint positions (crosses). }
\label{fig:2D_particle_results}
\end{figure}
%
Performance was evaluated by testing on 100 independently generated scenarios, each with 500 observations. At each observation time, an estimate of the position and velocity is made by taking the average of values from the array of particles. The resulting mean absolute error over all scenarios is shown in Fig.~\ref{fig:tracking_rmse}, again demonstrating a small improvement in accuracy from the smoother. The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the filter-smoother. Mean TNEES is displayed in Fig.~\ref{fig:tracking_mtnees}. It shows that while the smoother does not match the efficiency of the filter, it is a significant improvement over the filter-smoother, which has a Mean TNEES of $1$ at almost all time instants, because there is only one unique particle in the set.
%
\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.9\columnwidth]{tracking_pos_rmse.pdf}} \\
\subfloat[]{\includegraphics[width=0.9\columnwidth]{tracking_vel_rmse.pdf}}
\caption{Position (a) and velocity (b) mean absolute error over 100 scenarios using filter, filter-smoother, and smoother.}
\label{fig:tracking_rmse}
\end{figure}
%
\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{tracking_mtnees.pdf}
\caption{Mean TNEES over 100 scenarios using filter, filter-smoother, and smoother. Optimum value: 0.73}
\label{fig:tracking_mtnees}
\end{figure}
%
Finally, in Fig.~\ref{fig:varying_num_pts} the variation in accuracy and efficiency of the smoothing particle approximation is demonstrated as the number of smoothing particles varies. As expected, the rate of improvement diminishes and becomes negligible as the number of smoothing particles exceeds the number of filtering particles.
%
\begin{figure}
\centering
\subfloat[]{\includegraphics[width=0.45\columnwidth]{pos_RMSE_vs_numpts.pdf}}
\subfloat[]{\includegraphics[width=0.45\columnwidth]{MTNEES_vs_numpts.pdf}}
\caption{Average position error (a) and Mean TNEES (b) over 100 scenarios with varying number of smoother particles and 50 filter particles.}
\label{fig:varying_num_pts}
\end{figure}

@@

\appendices

@@

%\newpage
\bibliographystyle{IEEEtran}
\bibliography{Cleanbib}

\input{IEEE-Bunch.tex}
\input{IEEE-Godsill.tex}
